---
title: "Feature Importance GxE"
author: "Ty Thomas"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      comment = TRUE, out.width="110%", 
                      cache = TRUE,
                      cache.path = "../../../bmo_ase_gxe_nogit/Feature_importance_cache/", 
                      fig.path = "../../../bmo_ase_gxe_nogit/Feature_importance_cache/Figures/",
                      results='asis')

library(tidyverse)
library(patchwork)
library(pander)
library(DT)
library(tidymodels)
library(glmnet)
library(vip)
library(doParallel)
library(broom)
library(knitr)
library(kableExtra)
library(themis)
library(Boruta)
library(ggpattern)

theme_set(theme_bw()) 
set.seed(27597)

select <- dplyr::select
mutate <- dplyr::mutate

source("Functions_Feature_Importance.R")
```




## Read Data

The code below reads the data in.

```{r}
# Read in Data
sv <- read_tsv("../data/sv_per_gene.tsv") |> select(-is_bg, -strand)
dap <- read_tsv("../data/dap_seq_per_gene.tsv") |> select(-gene_class, -is_bg, -prom_s, -prom_e, -strand) |> mutate(across(everything(), ~ replace_na(.x, 0)))


# Combine the two together and order the columns
sv_dap <- left_join(sv, dap, by=c("gene", "chr", "is_gxe")) |> select(-gene, -chr) |> 
  mutate(n_sv_total = n_sv_total, .before = n_small_indel)

dap1 <- dap |> select(-gene, -chr)

sv1 <- sv |> select(-gene, -chr) 

# Make all characters factors
sv_dap_clean <- sv_dap |> 
  mutate(
    is_gxe = as.factor(is_gxe),
    across(where(is.character) | where(is.logical), as.factor)
  )

```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

# {.tabset}


## Stats reference

### Modeling and Statistical Metrics Reference Guide

This reference table outlines the different modeling frameworks, the statistical metrics used to evaluate feature importance and standalone predictive power, and how to interpret their values.

```{r reference_table, echo=FALSE, message=FALSE, warning=FALSE}

# Set pander options for clean document rendering
panderOptions('table.alignment.default', 'left')
panderOptions('table.split.table', Inf) # Prevents table from wrapping/splitting

ref_data <- tibble(
  `Model / Analysis Type` = c(
    "Logistic Regression Baseline (GLM)",
    "LASSO Regression",
    "Support Vector Machine (SVM)",
    "Random Forest (Ranger)",
    "Boruta Feature Selection Wrapper",
    "Univariate Predictive Capacity",
    "Information Theory Analysis"
  ),
  `Metric Used` = c(
    "Absolute Z-Score (|z-statistic|)",
    "Coefficient Magnitude",
    "Permutation Drop in ROC AUC",
    "Native Permutation Score",
    "Mean Importance (Z-Score) & Normalized Hits (normHits)",
    "Univariate ROC AUC Score",
    "Mutual Information (MI) Score"
  ),
  `How to Read It` = c(
    "**High = Good** (More statistically significant)",
    "**High = Good** (Stronger direct predictive weight)",
    "**High = Good** (Feature is critical to boundary accuracy)",
    "**High = Good** (Feature is heavily utilized across trees)",
    "**Focus on Final Decision Column**<br>**meanImp:** Higher = more reliable signal.<br>**normHits:** % of runs it beat random shadow noise.",
    "**0.5 = Random Guessing**<br>**1.0 = Perfect Separation**",
    "**0.0 = Completely Independent**<br>**Higher = More Shared Info**"
  ),
  `Brief Definition` = c(
    "Measures how many standard deviations a feature's estimated coefficient is away from zero. It identifies standard parametric significance.",
    "The absolute value of the model's optimized weight. Features with coefficients shrunk exactly to zero are completely filtered out by the penalty.",
    "Calculated by shuffling a feature's values in out-of-fold data and measuring the resulting drop in model accuracy. A massive drop means the model relies heavily on that feature.",
    " Shuffles a feature's values internally using Out-of-Bag (OOB) data during tree construction to measure how much it reduces overall node impurity.",
    "Compares real feature importance against randomized 'shadow' features over multiple iterations. Standardizes raw importance into a Z-score (meanImp) and uses a Binomial test on hit rates (normHits) to categorize features.",
    "Evaluates a single feature's standalone capability to separate the target groups (e.g., GxE vs non-GxE) without interacting with any other variables.",
    "Measures the total amount of overlapping information shared between a predictor and the target variable, successfully capturing both linear and highly complex non-linear relationships."
  ),
  `Important Context & Nuances` = c(
    "Prone to breaking or yielding extreme values/NAs when features are highly correlated (multicollinearity).",
    "Automatically handles multicollinearity by choosing one dominant feature from a highly correlated group and discarding the rest.",
    "Model-agnostic and highly robust, but computationally expensive because it must physically re-predict data for every single variable shuffled.",
    "Natively copes with highly correlated features, but will spread the 'credit' across them, which can make individual scores look lower than they honestly are.",
    "Gives a definitive 'Confirmed' or 'Rejected' label. Highly resilient to noise, but a feature can be statistically 'Confirmed' with a low Z-score if its weak signal is perfectly consistent.",
    "Your baseline check. If all individual features hover near 0.50, the dataset lacks an obvious, straightforward standalone signal.",
    "Measured in bits. Excellent for checking if a feature has a non-linear relationship (like a U-shape) that standard correlations completely miss."
  )
)

# Render the table cleanly for HTML output
pander(ref_data, caption = "Summary of Modeling Methods and Feature Importance Metrics")

```


<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">


## Base Data Models {.tabset}

### Structural Variants

TL;DR:

- Between all 4 models, the each variable shows importance at least once. 
- `n_small_indel` is consistently strong agross all 4 models. 
- Model roc_auc between 0.56 and 0.60

```{r}
SV_Models_All<- run_all_mods(sv1, Group_title = "Structural Variants")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">


### DAP_seq

TL;DR:

- LASSO is the best model, roc_auc 0.59, while other models are between 0.54 and 0.56
- LASSO being the best identifies 3 very strong factors, `n_small_indel_b73spec`, `n_snp_b73_spec`, and `n_small_indel_no_peak`
- The models don't share the same "most important" features. The only shared feature in the top 6 of each model is `frac_diff`

```{r}
DAP_Models_All<- run_all_mods(dap1, Group_title = "Dap_seq")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### Combined features

TL;DR:

- LASSO is the best model, roc_auc 0.6, while other models are between 0.54 and 0.57
- models don't share features in common. 

```{r}
SV_DAP_Models_All<- run_all_mods(sv_dap, Group_title = "Combined Features")

```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">


## Individual Predictive Power

TL;DR:

- Looking at the first table, there are 9 factors with AUC > 0.550. 8 have mutual information score above 0. this is because `n_sv_total` and `n_SNP` are highly correlated.
- `n_small_indel` is the most predictive of `is_gxe`. 
- When modeling using only the "best" features, the models don't increase in accuracy. 

```{r}

sv_dap_clean <- clean_dataset(sv_dap)
predictive_signals <- calculate_predictive_metrics(sv_dap_clean, "is_gxe")

pander::pander(predictive_signals, caption = "Standalone Predictive Capacity per Feature for is_gxe")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### Features with individual AUC > 0.550


```{r}
# selecting the best AUC features
best_AUC <- sv_dap |> 
  select(is_gxe, n_small_indel, n_SNP, n_snp_total, n_small_indel_no_peak,  n_snp_shared , n_small_indel_total, n_snp_no_peak, frac_diff)

Best_AUC_Models<- run_all_mods(best_AUC, Group_title = "Individual AUC > 0.550")

```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">


## Summarizing Features: Feature Totals vs Individuals {.tabset}

This Section separates out columns that are derived from the sums of other columns (totals), and columns that are not (individuals) to remove any strong correlations that may impact models. 

```{r}
#separating fetures into totals and individuals
totals_cols<- c("n_large_indel_total" , "n_small_indel_total", "n_snp_total", "n_total", "n_diff", "n_small_indel_genotype_spec", "n_large_indel_genotype_spec", "n_snp_genotype_spec", "n_sv_total")

dap_T <- dap1 |> 
  select(any_of(totals_cols), is_gxe)

dap_I <- dap1 |> 
  select(-any_of(totals_cols))


sv_T <- sv1 |> 
  select(any_of(totals_cols), is_gxe)

sv_I <- sv1 |> 
  select(-any_of(totals_cols))


sv_dap_T <- sv_dap |> 
  select(any_of(totals_cols), is_gxe)

sv_dap_I <- sv_dap |> 
  select(-any_of(totals_cols))
```


<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### SV: Totals

```{r}
#SV_Models_T<- run_all_mods(sv_T, Group_title = "SV - Totals")
```

This model will not run as there is only one variable. 

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### SV: Individuals

- `n_small_indel` is consistantly the better of the factors. 

```{r}
SV_Models_I<- run_all_mods(sv_I, Group_title = "SV - Individuals")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### DAP: Totals

- other than LASSO, `n_snp_total` is top two. 
- `n_total` and `n_small_indel_total` are also of interest


```{r}
DAP_Models_T<- run_all_mods(dap_T, Group_title = "DAP - Totals")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">


### DAP: Individuals

- `n_small_indel_no_peak` and `frac_diff` are interesting

```{r}
DAP_Models_I<- run_all_mods(dap_I, Group_title = "DAP - Individuals")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### Combined: Totals

- `n_sv_total` is the best in all models. The first time having a consistant best

```{r}
SV_DAP_Models_T<- run_all_mods(sv_dap_T, Group_title = "comb - Totals")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">

### Combined: Individuals

`n_small_indel_no_peak`

```{r}
SV_DAP_Models_I<- run_all_mods(sv_dap_I, Group_title = "comb - Individuals")
```

<hr style="border: 1px solid #4e5d6c; margin-top: 40px; margin-bottom: 40px;">



## Boruta


```{r}
# ==============================================================================
# 1. INITIALIZATION AND BORUTA RUN
# ==============================================================================

# Standardize and format data types uniformly across variables (factors/numerics)
sv_dap <- clean_dataset(sv_dap)

# Execute the Boruta wrapper feature selection using 250 iterative random forest runs.
# mcAdj = TRUE strictly applies Bonferroni corrections for high-dimensional testing.
bo_try <- Boruta(is_gxe~., data=sv_dap, mcAdj = TRUE, maxRuns = 250)

# Generate a quick base-R boxplot visualizing real features vs. shadow attribute benchmarks
plot(bo_try)

# Generate a quick trace plot showing feature importance progression across the 250 runs
plotImpHistory(bo_try)

# Render an interactive HTML data table showing ranked attribute importance metrics
DT::datatable(attStats(bo_try) |> arrange(-meanImp))


# ==============================================================================
# 2. DATA EXTRACTION AND RESTRUCTURING FOR CUSTOM PLOTTING
# ==============================================================================

# Extract summary stats, rank by mean importance, and convert row names to a dedicated column
bo_tab <- attStats(bo_try) |> arrange(-meanImp) |> 
  mutate(Features = rownames(attStats(bo_try) |> arrange(-meanImp)))

# Extract the full numerical history of Z-scores across all iterations into a dataframe
bo_hist <- bo_try[["ImpHistory"]] |> as.data.frame() 

# Pivot the history from a wide layout into a tidy long format for ggplot compatibility
bo_hist2 <- bo_hist |> pivot_longer(cols = colnames(bo_hist), names_to = "Features", values_to = "Importance")

# Join historical run data with categorical summary decisions, omitting benchmark shadow features
bo_all <- bo_hist2 |> 
  left_join(bo_tab, by = "Features") |> 
  filter(!Features %in% c("shadowMax", "shadowMean", "shadowMin")) 
  

# ==============================================================================
# 3. GEOM BOXPLOT & BARPLOT VISUALIZATIONS
# ==============================================================================

# Plot A: Vertical Boxplots tracking full Z-score distribution across iterations per feature
ggplot(bo_all, aes(x = fct_reorder(Features, -meanImp), y = Importance, fill = decision))+
         geom_boxplot(size = .3)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))+
  labs(y = "Importance", x = "Features", title = "Importance as calcualted from Boruta") +
  scale_fill_manual(values = c("#b3b3b3", "#b3b3ff", "#ffb3b3"))

# Plot B: Horizontal Bar Chart tracking static mean importance scores per feature
ggplot(bo_all, aes(y = fct_reorder(Features, -meanImp), x = meanImp, fill = decision))+
         geom_col()+
  theme_classic()+
  labs(y = "Importance", x = "Features", title = "Importance as calcualted from Boruta") +
  scale_fill_manual(values = c("#b3b3b3", "#b3b3ff", "#ffb3b3"))


# ==============================================================================
# 4. DOWNSTREAM MACHINE LEARNING FILTERING & MODELLING
# ==============================================================================

# Isolate the subset of attributes formally marked as "Confirmed" by the Boruta algorithm
best_ones <- attStats(bo_try) |> 
  filter(decision == "Confirmed")

# Subset the clean data matrix to retain only the target variable and Boruta-confirmed features
sv_dap1 <- sv_dap |> 
  select(all_of(c("is_gxe", rownames(best_ones))))

# Train, cross-validate, and score the final 4 machine learning pipelines using the refined feature subset
SV_DAP_Models_Boruta <- run_all_mods(sv_dap1, Group_title = "Boruta")
```


# Summary Plots

```{r}
# ==============================================================================
# 1. CATEGORIZE AND BIN UNIVARIATE METRICS
# ==============================================================================

# Create discrete thresholds for the Univariate Area Under the Curve (AUC) scores
predictive_signals1 <- predictive_signals |> 
  mutate(
    # Simplify the data source origins into descriptive category titles
    Type = if_else(Source == "SV Only", "Structural Variant", "DAP-seq"),
    
    # Bin continuous AUC values into 3 distinct, publication-ready tiers
    AUC = case_when(
      Univariate_AUC >= 0.550 ~ "≥0.550",
      Univariate_AUC < 0.550 & Univariate_AUC >= 0.525 ~ "0.525 to 0.550",
      TRUE ~ "<0.525"
    ),
    
    # Standardize the feature name variable handle for downstream merging
    Features = Variable
  )


# ==============================================================================
# 2. COMBINE UNIVARIATE DATA WITH BORUTA STATS
# ==============================================================================

# Merge the univariate screening dataframe with the Boruta importance summaries
predictive_signals2 <- predictive_signals1 |> 
  full_join(bo_tab, by = "Features") |> 
  
  # Keep only informative variables and drop non-specific aggregate features
  filter(
    Mutual_Information > 0,
    !Features %in% c("n_total", "n_shared")
  ) |> 
  
  # Duplicate the Boruta selection categorical column into a cleanly labeled header
  mutate("Boruta Decision" = decision)


# ==============================================================================
# 3. PLOT 1: STANDARD GGPLOT HORIZONTAL BAR CHART (UNIVARIATE POWER)
# ==============================================================================

# Plot features ordered by individual AUC performance, colored by discrete thresholds
ggplot(predictive_signals2, aes( y = fct_reorder(Variable, Univariate_AUC), x = Univariate_AUC, fill = fct_relevel(AUC,c("≥0.550", "0.525 to 0.550", "<0.525")))) +
         geom_col()+
  scale_fill_manual(values = c("#b3b3ff", "#b3b3b3", "#ffb3b3"))+
  theme_classic()+
  labs(x = "Univariate AUC", y = "Features", title = "Individual Predictive Power", fill = "AUC")+
  geom_hline(yintercept = 0.5) # Reference line mapping a zero-predictive-value threshold


# ==============================================================================
# 4. PLOT 2: GGPATTERN GEOM (DUAL ENCODING IMPORTANCE & SELECTION STATUS)
# ==============================================================================

# Construct a high-density publication figure mapping mean importance scores (X-axis),
# individual AUC categories (colors), and Boruta framework decisions (shading patterns)
p1 <- ggplot(predictive_signals2, aes(
    y = fct_reorder(Features, meanImp), 
    x = meanImp, 
    fill = fct_relevel(AUC, c("≥0.550", "0.525 to 0.550", "<0.525")),
    pattern = fct_relevel(decision, c("Confirmed", "Tentative", "Rejected")) 
  )) +
  geom_col_pattern(
    pattern_color = "grey95",       # Color tint of the fill pattern overlays
    pattern_fill = "grey95",        
    pattern_spacing = 0.02,         # Distance layout metrics separating lines/dots
    pattern_density = 0.1          
  ) +
  theme_classic() + 
  labs(
    x = "Importance", 
    y = "Features", 
    title = "Feature Importance",
    subtitle = "Based on Baruta and Individual Predictive Power (IPP)", # Note: "Baruta" typo here in title string
    fill = "AUC From IPP", 
    pattern = "Boruta Decision"       
  ) +
  scale_fill_manual(values = c("#b3b3ff", "#b3b3b3", "#ffb3b3")) +
  scale_pattern_manual(values = c("none", "stripe", "circle")) + # Maps unique overlays to selection states
  guides(fill = guide_legend(override.aes = list(pattern = "none"))) # Suppresses pattern clutter in the color legend

p1
# Export the customized dual-encoded figure to a local high-resolution file
ggsave("../../../bmo_ase_gxe_nogit/Features_Boruta_AUC.png", p1)
```


# Summary Tables

```{r}
gxe_features_ref <- tibble(
  `Feature Name` = c(
    "n_snp_total",
    "n_snp_no_peak",
    "n_SNP",
    "n_small_indel_no_peak",
    "n_small_indel",
    "n_small_indel_total",
    "n_snp_shared",
    "frac_diff"
  ),
  `Extended Feature Name` = c(
    "Total Promoter SNPs (DAP Consolidated)",
    "Promoter SNPs with No Peak Overlap",
    "Total Baseline Promoter SNPs",
    "Promoter Small Indels with No Peak Overlap",
    "Total Baseline Promoter Small Indels",
    "Total Promoter Small Indels (DAP Consolidated)",
    "Promoter SNPs Overlapping Shared Peaks",
    "Fraction of Genotype-Specific DAP Peaks"
  ),
  `Data Type` = c(
    "DAP_seq",
    "DAP_seq",
    "Structural Variants",
    "DAP_seq",
    "Structural Variants",
    "DAP_seq",
    "DAP_seq",
    "DAP_seq"
  ),
  `Boruta Importance` = c(
    0.770, 
    0.740, 
    0.703, 
    0.689, 
    0.647, 
    0.646, 
    0.635, 
    0.499
  ),
  `Univariate AUC` = c(
    0.573, 
    0.551, 
    0.575, 
    0.569, 
    0.591, 
    0.554, 
    0.558, 
    0.551
  ),
  `Mutual Information` = c(
    0.0006, 
    0.0002, 
    0.0002, 
    0.0002, 
    0.0010, 
    0.0003, 
    0.0004, 
    0.0011
  ),
  `Biological Description` = c(
    "The total count of Single Nucleotide Polymorphisms (SNPs) located within the gene's promoter region (+/-500bp from the Transcription Start Site), mapped within the DAP-seq consolidated framework.",
    "The count of promoter SNPs that do not physically overlap with any documented DAP-seq transcription factor binding peak, representing variants in non-binding promoter regions.",
    "The total baseline count of B73/Mo17 SNPs detected within the gene's promoter region, capturing sequence divergence between the parent genotypes.",
    "The count of small insertions or deletions (< 50 bp) in the promoter region that do not overlap with any DAP-seq transcription factor binding peak.",
    "The total baseline count of small insertions or deletions (< 50 bp) detected within the gene's promoter region.",
    "The total count of small insertions or deletions (< 50 bp) in the promoter region, tracked within the consolidated DAP-seq overlap architecture.",
    "The count of promoter SNPs that physically overlap with shared DAP-seq peaks (transcription factor binding sites present in both the B73 and Mo17 genotypes).",
    "The proportion of total transcription factor binding peaks in the promoter region that are genotype-specific (either B73-only or Mo17-only), serving as an index of localized regulatory divergence."
  )
)

# Preview the complete table
write_tsv(gxe_features_ref, file = "../../../bmo_ase_gxe_nogit/GxE_feature_importance.tsv", quote = "all")

pander(gxe_features_ref)
```



```{r}
# Create the consolidated data frame with all three conditions
model_comparison <- tibble(
  Model = c("LASSO", "Logistic_Regression", "Random_Forest", "SVM"),
  All_Features_AUC = c(0.5967, 0.5774, 0.5708, 0.5438),
  Boruta_Confirmed_Features_AUC = c(0.6028, 0.5834, 0.5706, 0.5469),
  Univariate_Filtered_Features_AUC = c(0.5924, 0.5825, 0.5451, 0.5672)
)

write_tsv(model_comparison, file = "../../../bmo_ase_gxe_nogit/GxE_prediction_models_AUC.tsv")

pander(model_comparison)
```



# Session info

```{r}
sessionInfo()
```

